---
title: "Sample Plots"
author: "Taylor Mackay"
output: pdf_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(fig.width=4, fig.height=3, fig.align = "center") 

rm(list = ls())

library(ggplot2)

```

\newpage

\tableofcontents

\newpage

# Introduction

This file contains examples of basic plots created using the ```ggplot2``` package in ``R`` and the corresponding code required to create each plot. All examples below require loading ```ggplot2```-- any other required packages are noted as needed in the included code.

__NOTE:__ The specific style of the plots below is specified by using ```theme_bcg``` in addition to the other plot options. This calls the code below in order to specify the plot style, font type and size, and center plot titles.

```{r theme}

# Setting options for plot formatting, including font type + size, and title
# alignment, using `minimal` theme

theme_bcg <- theme_minimal(base_size = 9, base_family = "Palatino") + 
  theme(plot.title = element_text(hjust = 0.5))

```


## Useful Resources

Useful websites with more information on ```R``` and ```ggplot2``` (click bulleted items for link to URL).  

* [RStudio ```ggplot2``` Cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
    + Two page PDF cheat sheet covering the basics of the ```ggplot2``` package
* [Gallery of ```ggplot2``` Examples](http://r-statistics.co/Top50-Ggplot2-Visualizations-MasterList-R-Code.html)
    + 50 different examples of plots, covering a range of plot types and customizations to things like legends and annotations
* [```R``` Datasets Package](http://stat.ethz.ch/R-manual/R-devel/library/datasets/html/00Index.html)
    + A list of the sample datasets available with ```R``` that are used in this document. Includes a detailed description of all variables in each dataset.

\newpage

# Univariate Plots

## Density Plot

```{r density.plot, fig.width=5, fig.height=4}

# Diamonds samples dataset has prices and other attributes of ~54,000 diamonds.

data(diamonds)

# Generating density plot 

ggplot(data = diamonds, aes(x = depth)) + geom_density() + 
  labs(title = "Sample Density Plot", 
       y = "Density", 
       x = "Depth") + 
  theme_bcg

```

\pagebreak

## Histograms with Grid Arrange

```{r histogram, fig.width=6, fig.height=3, message = F}

# `gridExtra` allows you to print multiple plots together 

library(gridExtra)

# Airquality sample dataset has measurements of temperature, windspeed, and 
# daily air quality in New York from May to September, 1973.

data("airquality")

# Default Histogram

p.1 <- ggplot(data = airquality) + geom_histogram(aes(x = Wind)) + 
  labs(title = "Default Histogram", 
       y = "Count", 
       x = "Average Daily Wind Speed (mph)") + 
  theme_bcg

p.2 <- ggplot(data = airquality) + geom_histogram(aes(x = Wind), binwidth = 2) +
  labs(title = "Bin Width Specified", 
       y = "", 
       x = "Average Daily Wind Speed (mph)") + 
  theme_bcg

# Using `grid.arrange` to print both plots side by side (by setting nrow = 1)

grid.arrange(p.1, p.2, nrow = 1)

```

\newpage

# Two-Way Plots

## Scatter Plot

```{r scatter.plot}

# Trees sample dataset has measurements of the height, weight, and length of
# 31 observations.

data(trees)

# Scatter Plot

ggplot(data = trees) + geom_point(aes(x = Height, y = Volume)) + 
  labs(title = "Sample Scatter Plot", 
       y = "Volume", 
       x = "Height") + 
  theme_bcg

```

\newpage

## Line Plot with Outcome Grouped by Factor Variable

```{r line.factor.plot, message = F}

# Orange sample data set has 7 measurements of age and circumference for 5 
# different oranges (total of 35 observations)

data(Orange)

# Start by creating a observation count by ID variable using `dplyr`. Note that 
# data needs to be in *long* form. 

library(dplyr)

df <- group_by(Orange, Tree) %>% 
  mutate(count = row_number()) 

# Creating re-ordered `tree` factor variable

df$Tree <- factor(df$Tree, levels = c(1,2,3,4,5))

# Line Plot-- notice options for setting x-axis ticks + legend label

ggplot(data = df) + geom_line(aes(x = count, y = circumference, 
                                  color = Tree)) +
  labs(title = "Sample Line Plot with Factor Groupings", 
       y = "Circumference (mm)", x = "Observation",
       color = "Tree") + 
  scale_x_continuous(breaks=seq(1, 7, 1)) + 
  theme_bcg

```

